---
title: "Stroke Prediction Data Analysis"
author: "Grace Lippert"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: yes
    toc_collapsed: yes
    code_folding: hide
    code_download: yes
    smooth_scroll: yes
    theme: lumen
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 3
    fig_height: 3
  word_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
editor_options: 
  chunk_output_type: inline
---

```{css, echo = FALSE}
#TOC::before {
  content: "Table of Contents";
  font-weight: bold;
  font-size: 1.2em;
  display: block;
  color: navy;
  margin-bottom: 10px;
}


div#TOC li {     /* table of content  */
    list-style:upper-roman;
    background-image:none;
    background-repeat:none;
    background-position:0;
}

h1.title {    /* level 1 header of title  */
  font-size: 22px;
  font-weight: bold;
  color: DarkRed;
  text-align: center;
  font-family: "Gill Sans", sans-serif;
}

h4.author { /* Header 4 - and the author and data headers use this too  */
  font-size: 15px;
  font-weight: bold;
  font-family: system-ui;
  color: navy;
  text-align: center;
}

h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Gill Sans", sans-serif;
  color: DarkBlue;
  text-align: center;
}

h1 { /* Header 1 - and the author and data headers use this too  */
    font-size: 20px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}

h2 { /* Header 2 - and the author and data headers use this too  */
    font-size: 18px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 16px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 14px;
  font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

/* Add dots after numbered headers */
.header-section-number::after {
  content: ".";

body { background-color:white; }

.highlightme { background-color:yellow; }

p { background-color:white; }

}
```
```{r setup, include=FALSE}
# Load required libraries
library(tidyverse)  # For data manipulation and visualization
library(ggplot2)    # For advanced plotting
library(naniar)     # For missing data visualization
library(skimr)      # For compact data summaries
library(GGally)     # For parallel coordinate plots and scatterplot matrices
library(plotly)     # For interactive visualizations
library(DT)         # For interactive tables
library(dplyr)
library(knitr)
library(MASS)

# Global chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)

# Custom theme for consistent visual design
theme_storytelling <- theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", margin = margin(b = 10)),
    plot.subtitle = element_text(size = 12, color = "grey40"),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    legend.position = "bottom",
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 20, 20, 20)
  )
```


# Introduction

According to the World Health Organization (WHO), in 2021 strokes were the third leading cause of death and disability globally, with an estimated 93.8 million cases.  Therefore, being able to quickly identify and predict stroke risk is vitally important.  For this analysis, we will be using a dataset with 5110 observations and 12 variables that contains information on whether or not a patient had a stroke along with potentially predictive demographic and health information.  The variables within this dataset include a patient's gender, age, bmi, type of employment, type of residence, and smoking status along with whether or not a patient has hypertension, heart disease, or has ever been married.  Finally, the target variable for this dataset is '[stroke]' which indicates whether or not a patient has had a stroke.

**Goals:** 

1. Identify variables within the dataset that can be useful in predicting a patient's risk of having a stroke.
2. Create a model that can be used to estimate a patient's predicted risk of a stroke using logistic regression. 

## Variable Descriptions

A complete list of the variables within the dataset is located below.  Note that the target variable for this analysis is '[stroke]' which indicates whether or not a patient has had a stroke.

```{r}
data_dictionary <- tribble(
  ~Variable_Name,         ~Description,                                                        ~Variable_Type,
  "[id]",                 "[Unique Identifier]",                                               "[Numeric]",
  "[gender]",             "[Gender of the patient (Male, Female, Other)]",                     "[Categorical]",
  "[age]",                "[Age of the patient (years)]",                                      "[Numeric]",
  "[hypertension]",       "[Whether the patient has hypertension (0 = No, 1 = Yes)]",          "[Numeric (Binary)]",
  "[heart_disease]",      "[Whether the patient has heart disease (0 = No, 1 = Yes)]",         "[Numeric (Binary)]",
  "[ever_married]",       "[Whether the patient was ever married (No, Yes)]",                  "[Categorical]",  
  "[work_type]",          "[Type of occupation (Private, Govt_job, Never_worked, etc.)]",      "[Categorical]",
  "[Residence_type]",     "[Type of residence (Rural, Urban)]",                                "[Categorical]",
  "[avg_glucose_level]",  "[Average glucose level in blood(mg/dL)]",                           "[Numeric]",
  "[bmi]",                "[Body mass index of the patient]",                                  "[Numeric]",
  "[smoking_status]",     "[Smoking status (smokes, never smoked, etc)]",                      "[Categorical]",
  "[stroke]",             "[If the patient had a stroke or not (0 = No, 1 = Yes)]",            "[Numeric (Binary)]",
 
)

datatable(
  data_dictionary,
  caption = "Table 1: Variable definitions and types for the Stroke Prediction Dataset",
  rownames = FALSE,
  options = list(pageLength = 10, dom = 'tip'))
```

# Data Quality, Distribution, and Relationship Assessment

## Missing Values and Reformatting 

Before moving on to further analyses, we will first consider the possibility of missing values and reformat several variables for ease of interpretability.  The variable '[bmi]' contains 201 missing values (imputed as 'N/A' in the original dataset).  Considering that 201 observations is only approximately 3.9% of of the overall 5110 observations we will simply remove these observations from the dataset.  

Furthermore, we will reformat several of the binary numeric variables so that they are categorical.  For instance '[hypertension]', '[heart_disease]', and '[stroke]' are all coded as binary numeric variable where 0 = 'No' and 1 = 'Yes').  Therefore we will recode them into categorical variables with the categories 'Yes' and 'No'.  Note that while '[hypertension]' and '[heart_disease]' were recoded directly, '[stroke]' was recoded into a new variable called '[stroke_recoded]' which preserves the original variable.  

```{r warning=FALSE}
strokepredict <- read.csv("https://pengdsci.github.io/datasets/bin-StrokePred/healthcare-dataset-stroke-data.csv")

strokepredict$hypertension <- factor(strokepredict$hypertension, levels = c(0,1), labels = c("No", "Yes"))

strokepredict$heart_disease <- factor(strokepredict$heart_disease, levels = c(0,1), labels = c("No", "Yes"))

strokepredict$stroke_recoded <- factor(strokepredict$stroke, levels = c(0,1), labels = c("No", "Yes"))

strokepredict$bmi <- as.numeric(strokepredict$bmi)

strokepredict <- na.omit(strokepredict)
```

## Distribution Analysis

We will begin by examining the distributions of the numeric variables in the dataset.  Understanding the shape, spread, and skewness of these variables can help us identify unusual patterns and outliers that may need to be taken into account when considering future modeling.

```{r}
hist_with_density <- function(data,             
                              variable,         
                              title = NULL,     
                              binwidth = NULL,  
                              color = "skyblue") {
  values <- data[[variable]]
  values <- values[!is.na(values)]
  dens <- density(values)
  if (is.null(binwidth)) {
    binwidth <- diff(range(values)) / 30
  }
  plot_ly() %>%
    add_trace(x = values, 
              type = "histogram",
              histnorm = "probability density",
              marker = list(color = color,
                            line = list(color = "white", 
                                        width = 1)),
              opacity = 0.7,
              name = "Histogram",
              nbinsx = 30) %>%
    add_trace(x = dens$x, 
              y = dens$y,
              type = "scatter", 
              mode = "lines",
              line = list(color = "darkred", 
                          width = 3),
              fill = "tozeroy",
              fillcolor = "rgba(255,0,0,0.1)",
              name = "Density Curve") %>%
    layout(title = ifelse(is.null(title), 
                         paste("Distribution of", variable), title),
           xaxis = list(title = list(text= variable,
                                     standoff = 0),  
                        titlefont = list(size = 14)
                        ),
           yaxis = list(title = "Density",
                        titlefont = list(size = 14)),
           legend = list(x = 0.85, 
                         y = 0.95),
           hovermode = "closest",
           margin = list(l=50, r = 20, t = 100, b = 70))
}

p1 <- hist_with_density(strokepredict, "age", "Age Distribution")
p2 <- hist_with_density(strokepredict, "avg_glucose_level", "Average Glucose Level Distribution")
p3 <- hist_with_density(strokepredict, "bmi", "BMI Distribution")

fig1 <- subplot(p1, p2, p3, 
               nrows = 1, 
               margin = 0.06,
               shareX = FALSE, 
               shareY = FALSE,
               titleX = TRUE, 
               titleY = TRUE)

fig1 <- fig1 %>% layout(title = "Distributions of Continous Variables", showlegend = FALSE)

fig1
```

Age has a multimodal distibution that is much more evenly spread out, indicating that the age of patients within this dataset varies widely.  Alternatively, while also being a multimodel distribution (bimodal), average glucose level has a noticeable right skew, indicating that most average glucose levels are lower with some outliers that are much higher.  According to the CDC diabetes testing guidelines, a normal glucose level is below 117 mg/dL, a prediabetes glucose level is between 117 mg/dL and 154 mg/dL, and a diabetes glucose level is above 154 mg/dL.  As we can see in the above distribution, most values appear to be in the normal-prediabates range with some skewing much higher into the diabetes range.  BMI has a unimodal right-skewed distribution.  This indicates that there are some outlier values that indicate higher BMIs.  According to the CDC, a BMI of below 18.5 is considered underweight, 18.5-24.9 is a healthy weight, 25.0-29.9 is overweight, and 30.0 or higher is considered obese.  In the distribution above, we can see that the distribution is centered around 10.0-50.0, with some values being notably higher than this range.    

Next we will consider the frequency distribution of our categorical variables.  Examining these distributions will allow us to identify any categories that contain too few values to meaningfully analyze.  

```{r}
cat_plot_sorted <- function(data, var, title=NULL, color="#1f77b4", include_na = TRUE){
  tab <- as.data.frame(table(data[[var]], useNA=ifelse(include_na, "ifany", "no")))
  colnames(tab) <- c("category", "count")
  tab$count <- as.numeric(tab$count)
  tab <- tab[!is.na(tab$category), ]
   if(nrow(tab) == 0){
    stop(paste("No valid data for variable:", var))
  }

  tab <- tab[order(tab$count, decreasing = TRUE), ]
  tab$category <- factor(tab$category, levels = tab$category)
  plot_ly() %>%
    add_bars(
      x = tab$category,
      y = tab$count,
      text = tab$count,
      textposition = "auto",
      marker = list(color = color)
    ) %>%
    layout(
      title = ifelse(is.null(title), var, title),
      xaxis = list(title = var),
      yaxis = list(title = "Count"),
      margin = list(b = 100)
    )
}

cat1 <- cat_plot_sorted(strokepredict, "gender", "Gender Distribution")
cat2 <- cat_plot_sorted(strokepredict, "hypertension", "Hypertension Distribution")
cat3 <- cat_plot_sorted(strokepredict, "heart_disease", "Heart Disease Distribution")
cat4 <- cat_plot_sorted(strokepredict, "ever_married", "Ever Married Distribution")
cat5 <- cat_plot_sorted(strokepredict, "work_type", "Work Type Distribution")
cat6 <- cat_plot_sorted(strokepredict, "Residence_type", "Residence Type Distribution")
cat7 <- cat_plot_sorted(strokepredict, "smoking_status", "Smoking Status Distribution")
cat8 <- cat_plot_sorted(strokepredict, "stroke_recoded", "Stroke Distribution")

fig2 <- subplot(list(cat1, cat2, cat3, cat4, cat5, cat6, cat7, cat8),
               nrows = 2,
               shareX = FALSE,
               titleX = TRUE,
               titleY = TRUE,
               margin = 0.08) %>%
  layout(title = "Distribution of Categorical Variables",
         showlegend = FALSE,
         font = list(size = 12)
         )

fig2

```

As we can see in the barplot for '[gender]' above, there is only one observation within the 'Other' category of the '[gender]' variable.  Given that this is only one observation that can't be meaningfully combined with one of the other two categories, we will remove this category since it is too small to be useful for analysis.

```{r}
strokepredict <- strokepredict %>% filter(gender != "Other")
```

## Correlation between Numeric Variables

Finally, before assessing the relationship between '[stroke]' and our potential predictive variables, we will briefly consider the correlations between our numeric variables.  The correlations between '[age]', '[avg_glucose_level]', and '[bmi]' are represented in the figure below:

```{r, warning=FALSE}
pairwise_scatter_matrix <- function(data, vars) {
  n_vars <- length(vars)
  subplots_list <- list()
  for (i in 1:n_vars) {
    for (j in 1:n_vars) {
      var_x <- vars[j]  
      var_y <- vars[i]  
      if (i == j) {
        dens <- density(data[[var_y]], na.rm = TRUE)
        dens_df <- data.frame(x = dens$x, y = dens$y)
        p <- plot_ly(dens_df, x = ~x, y = ~y, 
                     type = 'scatter', 
                     mode = 'lines',
                     fill = 'tozeroy',
                     line = list(color = '#1f77b4', width = 2),
                     fillcolor = 'rgba(31, 119, 180, 0.3)',
                     showlegend = FALSE) %>%
          layout(xaxis = list(title = var_y, showgrid = TRUE),
                 yaxis = list(title = "Density", showgrid = TRUE))
      } else if (i > j) {
        valid_data <- data[complete.cases(data[c(var_x, var_y)]), c(var_x, var_y)]
        if (nrow(valid_data) > 5) {
          loess_fit <- loess(valid_data[[var_y]] ~ valid_data[[var_x]])
          loess_pred <- predict(loess_fit, 
                                data.frame(x = valid_data[[var_x]]))
          loess_df <- data.frame(x = valid_data[[var_x]], 
                                 y = loess_pred)
        }
        p <- plot_ly() %>%
          add_trace(x = valid_data[[var_x]], 
                   y = valid_data[[var_y]],
                   type = 'scatter', mode = 'markers',
                   marker = list(size = 5, color = 'steelblue', 
                               opacity = 0.6),
                   name = 'Data', showlegend = FALSE)
        if (exists("loess_df") && nrow(loess_df) > 0) {
          loess_df <- loess_df[order(loess_df$x),]
          p <- p %>% add_trace(x = ~x, y = ~y,
                              data = loess_df,
                              type = 'scatter', mode = 'lines',
                              line = list(color = '#d62728', width = 2),
                              name = 'Loess', showlegend = FALSE)
        }
        
        p <- p %>% layout(xaxis = list(title = var_x, 
                                       showgrid = TRUE,
                                        standoff = 0), 
                         yaxis = list(title = var_y, showgrid = TRUE))
        
      } else {
        cor_value <- cor(data[[var_x]], data[[var_y]], use = "complete.obs")
        cor_text <- sprintf(" %.3f", cor_value)
        p_value <- cor.test(data[[var_x]], data[[var_y]], 
                           use = "complete.obs")$p.value
        pval <- round(p_value, 3)
        cor_text <- paste0(cor_text, "\n p =", pval)
        p <- plot_ly( x = 1, y = 1, type = "scatter", mode = "markers", marker = list(opacity = 0)) %>%
          layout(
           xaxis = list(
              title = "",                    
              showticklabels = FALSE,        
              showgrid = FALSE,              
              zeroline = FALSE,              
              showline = FALSE,              
              showaxeslabels = FALSE,        
              ticks = "",                    
              ticklen = 0,                   
              tickwidth = 0                  
            ),
            yaxis = list(
              title = "",                    
              showticklabels = FALSE,        
              showgrid = FALSE,              
              zeroline = FALSE,              
              showline = FALSE,              
              showaxeslabels = FALSE,        
              ticks = "",                    
              ticklen = 0,                   
              tickwidth = 0                  
            ),
            plot_bgcolor = 'white',           
            paper_bgcolor = 'white',
            annotations = list(
              text = cor_text,
              font = list(size = 18, color = '#7f7f7f'),
              xref = "paper", 
              yref = "paper",
              x = 0.5, 
              y = 0.5,
              showarrow = FALSE
            )
          )
      }
      
      subplots_list <- append(subplots_list, list(p))
    }
  }
  fig <- subplot(subplots_list, 
                 nrows = n_vars, 
                 shareX = FALSE, 
                 shareY = FALSE,
                 titleX = TRUE, 
                 titleY = TRUE,
                 margin = 0.05)
  fig <- fig %>% layout(title = "Pairwise Correlation Between Continuous Variables",
                        showlegend = FALSE,
                        plot_bgcolor = 'white',
                        paper_bgcolor = 'white')
  
  return(fig)
}

strokevars <- c("age", "avg_glucose_level", "bmi")

fig3 <- pairwise_scatter_matrix(strokepredict, strokevars)

fig3
```

All of the numeric variables exhibit weak correlations, which indicates that the variables measure distrinct demographic and health-related features.  The highest correlation was between gender and BMI at $r = 0.333$.  However, this correlation is not high enough to indicate collinearity between these two variables.

# Variable Contributions to Stroke Risk

In order to determine which variables are most associated with high stroke risk, we will begin with an initial visual comparison of each variable, comparing patients who have had a stroke and who have not.  This will help us determine which variables should be included in the candidate model.

## Grouped Boxplots - Continuous Variables

```{r}
create_boxplot <- function(data, variable, var_title) {
  # Filter out NA values
  plot_data <- data %>% filter(!is.na(.data[[variable]]))
  
  p <- plot_ly() %>%
    add_trace(
      data = plot_data %>% filter(stroke_recoded == "No"),
      x = ~stroke_recoded,
      y = as.formula(paste0("~", variable)),
      type = "box",
      name = "No Stroke",
      fillcolor = "skyblue",  
      line = list(color = "steelblue", width = 2),  
      marker = list(color = "orange", size = 3),  
      boxmean = "sd",
      legendgroup = "0",
      showlegend = TRUE,
      hovertemplate = paste0(
        '<b>No Stroke</b><br>',
        var_title, ': %{y:.1f}<br>',
        'Median: %{median:.1f}<br>',
        'Mean: %{mean:.1f}<extra></extra>'
      )
    ) %>%
    add_trace(
      data = plot_data %>% filter(stroke_recoded == "Yes"),
      x = ~stroke_recoded,
      y = as.formula(paste0("~", variable)),
      type = "box",
      name = "Stroke",
      fillcolor = "skyblue",  
      line = list(color = "steelblue", width = 2),  
      marker = list(color = "orange", size = 3),  
      boxmean = "sd",
      legendgroup = "1",
      showlegend = TRUE,
      hovertemplate = paste0(
        '<b>stroke</b><br>',
        var_title, ': %{y:.1f}<br>',
        'Median: %{median:.1f}<br>',
        'Mean: %{mean:.1f}<extra></extra>'
      )
    ) %>%
    layout(
      title = list(text = var_title, 
                   font = list(size = 14)),
      margin = list(l = 50, r = 50, b = 50, t = 70),
      xaxis = list(
        title = " ",
        ticktext = c("No Stroke", "Stroke"),
        tickvals = c(0, 1),
        tickfont = list(size = 12)
      ),
      yaxis = list(
        title = var_title,
        titlefont = list(size = 12),
        tickfont = list(size = 10)
      ),
      hovermode = "compare",
      plot_bgcolor = '#f8f9fa',
      paper_bgcolor = 'white',
      legend = list(
        title = list(text = "Stroke"),
        orientation = "h",
        yanchor = "bottom",
        y = 1.02,
        xanchor = "center",
        x = 0.5
      )
    )
  
  return(p)
}

box1 <- create_boxplot(data = strokepredict, variable="age", var_title="Age")
box2 <- create_boxplot(data = strokepredict, variable="avg_glucose_level", var_title="Average Glucose Level")
box3 <- create_boxplot(data = strokepredict, variable="bmi", var_title="BMI")

fig4 <- subplot(box1, box2, box3, 
               nrows = 1, 
               margin = 0.05,  
               shareX = FALSE, 
               shareY = FALSE,
               titleX = TRUE, 
               titleY = TRUE)

fig4 <- fig4 %>% layout(title = "Boxplots of Continous Variables (Stroke vs. No Stroke)",
                     showlegend = FALSE)
fig4


```

**Age** shows a strong visual distinction between the two categories.  The boxplot for patients who had had a stroke was much higher than the boxplot for those who had not, with both the mean and median for the stroke category being noticably higher.  This indicates that patients who had had a stroke were skewed older than patients who had not.  

**Average Glucose Level** also showed a strong visual distinction between the two categories.  Once again, the mean and median for patients who had had a stroke was much higher than for patients who had not.  This indicates that patients who had had a stroke tended to have a higher average glucose level than patients who had not.

**BMI** on the other hand did not show a strong visual distinction between the two categories.  While the 'no stroke' category had more outliers, this may very well be due to the 'no stroke' categories' notably higher sample size.  The lack of a strong visual difference does not mean that BMI is not a variable that would be useful in our predictive model (it very well could be), however it will not be added to our initial reduced model given its lack of visual distinction.

## Grouped Barplots - Categorical Variables

```{r}
create_grouped_bar_chart <- function(data, categorical_var, title = NULL) {
  
  summary_data <- data %>%
    group_by(!!sym(categorical_var), stroke_recoded) %>%
    summarise(count = n(), .groups = 'drop') %>%
    group_by(!!sym(categorical_var)) %>%
    mutate(
      percentage = count / sum(count) * 100,
      total = sum(count)
    ) %>%
    ungroup()
  summary_data <- summary_data %>%
    mutate(
      hover_text = paste0(
        "Variable: ", !!sym(categorical_var), "<br>",
        "Stroke: ", stroke_recoded, "<br>",
        "Count: ", count, "<br>",
        "Percentage: ", round(percentage, 1), "%"
      )
    )
  
  fig <- plot_ly(
    data = summary_data,
    x = ~get(categorical_var),
    y = ~percentage,
    color = ~stroke_recoded,
    type = "bar",
    colors = c("#1f77b4", "#ff7f0e"),
    text = ~paste("",round(percentage, 1),"%"),#count),
    textposition = "outside",
    hoverinfo = "text",
    hovertext = ~hover_text,
    marker = list(
      line = list(color = "white", width = 1.5)
    )
  ) %>%
    layout(
      title = ifelse(is.null(title), 
                     paste("Distribution of", categorical_var, "by Stroke Status"),
                     title),
      xaxis = list(
        title = title,
        standoff = 0
      ),
      margin = list(t=50),
      yaxis = list(
        title = "Percentage (%)",
        ticksuffix = "%"
      ),
      barmode = "group",
      legend = list(
        title = list(text = "Stroke"),
        orientation = "h",
        yanchor = "bottom",
        y = 0.9,
        xanchor = "right",
        x = 0.75
      ),
      plot_bgcolor = "#f5f5f5",
      paper_bgcolor = "white"
    )
  
  return(fig)
}

group1 <- create_grouped_bar_chart(strokepredict, categorical_var="gender", title = "Gender")
group2 <- create_grouped_bar_chart(strokepredict, categorical_var="hypertension", title = "Hypertension")
group3 <- create_grouped_bar_chart(strokepredict, categorical_var="heart_disease", title = "Heart Disease")
group4 <- create_grouped_bar_chart(strokepredict, categorical_var="ever_married", title = "Ever Married")
group5 <- create_grouped_bar_chart(strokepredict, categorical_var="work_type", title = "Work Type")
group6 <- create_grouped_bar_chart(strokepredict, categorical_var="Residence_type", title = "Residence Type")
group7 <- create_grouped_bar_chart(strokepredict, categorical_var="smoking_status", title = "Smoking Status")

fig5 <- subplot(group1, group4, group5, group6,
               nrows = 2, 
               margin = 0.09, 
               shareX = FALSE, 
               shareY = FALSE,
               titleX = TRUE, 
               titleY = TRUE)

fig5 <- fig5 %>% layout(title = "Barplots of Categorical Variables - Demographic (Stroke vs. No Stroke)",
                     showlegend = FALSE)
fig5

fig6 <- subplot(group2, group3, group7, 
               nrows = 2, 
               margin = 0.09, 
               shareX = FALSE, 
               shareY = FALSE,
               titleX = TRUE, 
               titleY = TRUE)

fig6 <- fig6 %>% layout(title = "Barplots of Categorical Variables - Health (Stroke vs. No Stroke)",
                     showlegend = FALSE)
fig6



```

The above barplots indicate that whether or not a patient had **Hypertension** or **Heart Disease** was visually strongly associated with whether or not they had had a stroke. Only 3.3% of patients who did not have hypertension had a stroke while 13.3% of patients who did have hypertension had a stroke.  Heart disease was even more strongly associated with having a stroke with only 3.6% of patients without heart disease having had a stroke and 16.5% with heart disease having had a stroke.  This indicates that both of these variables are highly associated with whether or not a patient had had a stroke and should be included in the candidate model.  Once again, this does not mean that the other variables would be useful in our predictive model, however they will not be added to our initial reduced model given their lack of visual distinction.

# Logistic Regression Analysis

## Brief Introduction to Logistic Regression

The analytical form of the logistic regression model is shown below (note that stroke = 1 indicates that the patient has a stroke):

$$
P[\text{Stroke}=1]=\frac{\exp(\beta_0+\beta_1R_1+\beta_2R_2+ ...+\beta_pR_p)}{1+\exp(\beta_0+\beta_1R_1+\beta_2R_2+ ...+\beta_pR_p}
$$

where $R_1,R_2,...,R_p$ are risk factors like age, average glucose level, hypertension, etc. and $\beta_0, \beta_1,...,\beta_p$ are the regression coefficients representing the effects of the predictors on the log-odds of a stroke occurring.  If a coefficient $\beta_j=0$ in a logistic regression model, it means that the corresponding predictor has no effect on the log-odds of the outcome (i.e. the patient does not have hypertension therefore that risk factor would not be relevant).

It should be noted that a positive regression coefficient ($\beta_j$) indicates that as the predictor increases the log-odds of the outcome increase.  Alternatively a negative regression coefficient indicates that as the predictor increases the log-odds of the outcome decrease.  

The odds ratio (OR) in logistic regression is a measure for interpreting the effect of a predictor on the outcome.  It is defined as:

$$
\text{OR}=e^{\beta_j}
$$
It represents how the odds of the outcome ($\text{Stroke}=1$) change when the predictor inceases by one unit, holding the other variable constant.

## Building the Logistic Regression Model

We will use our previous work up entil this point to develop two baseline models, a reduced model based on the highly associated variables we identified previously, and a full model which includes all potentially significant variables.  Then we will use forward selection to determine the optimal candidate model.  The results of this optimal candidate model are shown below:

```{r}
strokepredict$stroke <- as.factor(strokepredict$stroke)
strokepredict$gender <- as.factor(strokepredict$gender)
strokepredict$hypertension <- as.factor(strokepredict$hypertension)
strokepredict$heart_disease <- as.factor(strokepredict$heart_disease)
strokepredict$ever_married <- as.factor(strokepredict$ever_married)
strokepredict$work_type <- as.factor(strokepredict$work_type)
strokepredict$Residence_type <- as.factor(strokepredict$Residence_type)
strokepredict$smoking_status <- as.factor(strokepredict$smoking_status)

strokereg <- strokepredict %>% 
  dplyr::select(-stroke_recoded)


reduced_model <- glm(stroke ~ age + hypertension + heart_disease + avg_glucose_level, 
                     data = strokereg, 
                     family = binomial(link = "logit"))
#kable(summary(reduced_model)$coef, caption="Inferential table of the reduced logistic regression model")

full_model <- glm(stroke ~ ., 
                  data = strokereg, 
                  family = binomial(link = "logit"))
#kable(summary(full_model)$coef, caption ="Inferential table of the full logistic regression model")

forward_model <- stepAIC(reduced_model, 
                         scope = list(lower = reduced_model, upper = full_model),
                         direction = "forward", 
                         trace = FALSE)

kable(summary(forward_model)$coef, caption = "Inferential table of the optimal candidate logistic regression model")
```

The optimal logistic regression model identifies four statistically significant predictors of stroke risk, '[age]', '[hypertension]', '[heart_disease]', and '[avg_glucose_level]'.  The intercept and the coefficients were all statistically significant ($p<0.05$).  Furthermore, all of the regression coefficients were positive, indicating that higher values of age and/or average glucose level as well as the presence of hypertension and/or heart disease increase stroke risk.  Interestingly, this model is the same as our reduced model based off of our previous visual analysis.

# Conclusion

## Using the Model

Predicted stroke risk can be calculated using the following steps.  First compute the linear predictor ($\text{LP}$) using the following formula:

$$
\text{LP}=-7.6603+0.0675(\text{age})+0.5396(\text{hypertension})+0.4043(\text{heart_disease})+0.0048(\text{avg_glucose_level})
$$

Where each predictor is entered as follows:

- **age**: patient's age in years 
- **hypertension**: 0 if doesn't have hypertension, 1 if does have hypertension
- **heart_disease**: 0 if doesn't have heart disease, 1 if does have heart disease
- **avg_glucose_level**: patient's average glucose level (mg/dL)

Using the logistic regression model formula, the stroke risk score (probability of having a stroke) is calculated as follow:

$$
P[\text{Stroke=1}]=\frac{\exp(\text{LP})}{1+\exp(\text{LP})}
$$

## Recommendations

Based on the regression model, there are several recommendations that can be made.  Firstly, patients with hypertension should be prioritized for intensive blood pressure management.  Regular blood pressure monitoring, medication adherence, dietary modifications, and increased physical activity may help reduce stroke risk in this population.  Second, patients with elevated average glucose levels should receive additional screening and support for blood sugar management. Although the effect of glucose level is gradual, persistent elevations can substantially increase stroke risk over time. Interventions such as dietary counseling, weight management, and diabetes prevention or management programs may help mitigate this risk.  Third, individuals with a history of heart disease should be considered a higher-risk group for stroke and may benefit from closer cardiovascular monitoring and preventive care. Managing cardiovascular risk factors, maintaining prescribed medications, and attending regular follow-up appointments may help reduce the likelihood of future stroke events.  Finally, age was the strongest predictor in the model, with stroke risk increasing steadily as patients grow older. While age itself cannot be modified, healthcare providers can use age as a screening tool to identify individuals who may benefit from more frequent cardiovascular assessments, earlier intervention, and targeted preventive strategies. Older adults who also have hypertension, heart disease, or elevated glucose levels should be considered particularly high-risk and may warrant more intensive monitoring and prevention efforts.







